Software architectures to integrate workflow engines in science gateways

### 2016-10-13 paper.tex
Updates in figures
Still more answers to the reviewers

\usepackage[x11names,dvipsnames,table]{xcolor} %for use in color links
\usepackage[subrefformat=parens,labelformat=parens]{subfig}
\usepackage{svg}
\usepackage{url}
\usepackage{flushend}
\usepackage{xspace}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{multirow}
\usepackage{pbox}
\usepackage{pdfcomment}
\definecolor{ao(english)}{rgb}{0.0, 0.5, 0.0}
\newcommand{\todo}[2]{\pdfmargincomment[color=red,author=#1,open=true]{#2}}
%\newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
%\newcommand{\closednote}[4]{} % closed notes are hidden. This should obviously be a parameter of \note rather than a separate command.
%\newcommand{\answerednote}[4]{\pdfmargincomment[color=green,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}
%\newcommand{\closedanswerednote}[6]{}
\newcommand{\correction}[1]{\color{blue}#1\color{black}\xspace}

@@

\correction{The reminder of this section provides} background
information and definitions of workflow engines, science gateways and
infrastructures. In Section~\ref{sec:architectures}, we describe six
architectures that cover the main patterns that we encountered in the
various systems that we studied during the last few years. The
architectures are described within a consistent framework that
underlines the functional interactions between their main software
components. \correction{In Section~\ref{sec:metrics}, we introduce
  metrics that measure integration complexity, robustness,
  extensibility, scalability and \correction{functionality}. These
  metrics are specifically designed to measure the ability of the
  architectures to address workflow-related issues commonly
  encountered in science gateways. In Section~\ref{sec:evaluation}, we
  evaluate the different architectures based on the proposed metrics.}
We finally compare the architectures and highlight recommendations for
science gateway architects and developers.

@@

See Figure~\subref*{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a model in a
Ruby on Rails application\footnote{\correction{Note that the
    particular case of portlets can be a bit ambiguous: when portlets
    are deployed remotely, using a protocol such as Web Services for
    Remote Portlets (WSRP), they become independent software
    components according to our definition because they could be
    operated by different teams on different hardware systems. In this
    case, the system should be classified in the service invocation
    architecture since the portlets actually become Web
    Services. The same comment holds for remote object
    invocation.}}. The workflow engine and the science gateway usually
share a database where applications, users and other resources are
stored.  In this model, task and data control are both initiated from
the science gateway. Interactions \texttt{b} and \texttt{c$_2$} are
initiated from the workflow engine, while \texttt{c$_1$} comes from
other parts of the science gateway, for instance a data management
interface. As in any other model, the installation of new workflows in
the science gateway (\texttt{a$_2$}) and infrastructure
(\texttt{a$_1$}) is done by an administrator, for security
reasons. This is the model adopted in the Catania Science Gateway
Framework~\cite{ardizzone2012decide} (see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}), in the Distributed
application runtime environment
(DARE)~\cite{maddineni2012distributed}, in
Galaxy~\cite{goecks2010galaxy}, and in the LONI Pipeline
Environment~\cite{dinov2009efficient}. Note that tightly integrated
architectures may provide advanced workflow edition features which are
not covered by our analysis.
%The Catania Science Gateway Framework would be a good illustration here (workflow engine is a Java portlet), but I can't find a suitable figure about it.
%CIPRES~\cite{miller2010creating} (on which the Neuroscience gateway
%described in~\cite{sivagnanam2015early} is based) % \note{Tristan}{There is no workflow engine in there.}

@@

See Figure~\subref*{archi:sub-task}. The workflow engine is wrapped in
a particular task that can submit sub-tasks to the science gateway
through interaction \texttt{e}. The workflow engine keeps track of
the dependencies between the sub-tasks, but their execution is
delegated to the science gateway that executes them on the
infrastructure through interaction \texttt{b}. Although the science
gateway has no global vision of the workflow, it completely \correction{controls}
the sub-tasks. \correction{For instance, it may cancel} them when
the \correction{workflow} is canceled. The science gateway may also implement
mechanisms to facilitate the handling of task dependencies, for
instance basic dependency lists as available in most cluster managers
(see for instance attribute \texttt{depend} of option \texttt{-W} in
Torque\footnote{\url{http://docs.adaptivecomputing.com}}).

@@

\correction{\section{Architectural patterns in real systems}}

@@

\correction{This section illustrates the architectural patterns on a
  few real systems. The architecture diagrams presented are the
  original ones as published by the system developers, which we
  annotated using the interactions and components in
  Figure~\ref{fig:architectures}. }

@@

Figure~\ref{fig:vip-architecture} shows the architecture of the
Virtual Imaging Platform, where the MOTEUR workflow
engine~\cite{GLAT-08i} is invoked through a service.

\begin{figure*}
\centering
%\def\svgwidth{1.5\columnwidth}
\includegraphics[width=0.9\textwidth]{figures/VIP}
\caption{Architecture of the Virtual Imaging Platform (service
  invocation).  Step \texttt{3} corresponds to interaction \texttt{d}
  in Figure~\ref{fig:architectures}(II), step \texttt{2} corresponds to
  interaction \texttt{c$_1$}, step \texttt{4} maps to interaction
  \texttt{b}, and steps 7 and 9 map to \texttt{c$_2$} (in VIP, workflow data transfers are
  performed by the tasks and embedded in their descriptions).}
\label{fig:vip-architecture}
\end{figure*}

@@

Task encapsulation is implemented in CBRAIN~\cite{SHER-14} where it is
used to integrate the FSL toolkit~\cite{Jenkinson2012782} and the PSOM
workflow engine~\cite{bellec2012pipeline}. The CBRAIN-FSL integration
allows leveraging FSL pipelines written in low-level workflow
languages (Linux executables and scripts) that submit tasks uniformly
through a specific tool called fsl\_sub. This integration is shown in
Figure~\subref*{fig:cbrain-fsl-architecture}, where the science gateway is
represented by components \texttt{CBRAIN portal} and \texttt{CBRAIN
  execution server} and the infrastructure consists of \texttt{Storage
  servers} and \texttt{Computing cluster with shared file system}.

@@

\begin{figure*}
\centering
\subfloat[CBRAIN-FSL integration.]{
\def\svgwidth{\columnwidth}
\input{figures/CBRAIN-FSL.pdf_tex}
\label{fig:cbrain-fsl-architecture}
} \hfill \subfloat[CBRAIN-PSOM integration. Figure
reproduced from~\cite{GLAT-16}.]{\def\svgwidth{\columnwidth}
  \input{figures/CBRAIN-PSOM.pdf_tex
}
\label{fig:cbrain-psom-architecture}
}
\caption{CBRAIN architecture for workflow engine integration
  (task encapsulation).  1: User sends data and workflow execution request to
  storage server(s) and CBRAIN portal. 2: CBRAIN portal sends
  execution request to execution server on cluster. 3: Execution
  server transfers data from storage server(s). 4-5: Execution server
  starts workflow engine (FSL tool or PSOM master) via task
  scheduler. 6: Workflow engine submits sub-tasks to execution server
  (FSL tasks or PSOM agents). 7-8: Execution server starts sub-tasks
  through task scheduler (FSL sub-tasks or PSOM workers). FSL sub-tasks will run locally
  instead of being submitted again to CBRAIN through interaction 6. 8':
  (PSOM only) PSOM master and workers execute workflow. 9. Execution
  server transfers results to storage server(s). Interaction
  \texttt{b} in Figure~\ref{fig:architectures}(III) is implemented by steps
  \texttt{4}, \texttt{5}, \texttt{7} and \texttt{8} (regular
  interactions with batch manager). Interaction \texttt{c} is
  implemented by steps \texttt{3} and \texttt{9}. Interaction
  \texttt{e} is implemented by step \texttt{6} (for FSL: through a
  modified version of the fsl\_sub script available in
  \url{https://github.com/aces/cbrain-plugins-neuro}; for PSOM:
  through a specific development in PSOM).}
\label{fig:cbrain-sub-tasking}
\end{figure*}

The pool model was implemented in the SHIWA pool~\cite{ROGE-13}
diagrammed in Figure~\ref{fig:shiwa-pool-architecture}, where agents
can wrap different types of workflow engines to execute workflows
expressed with different languages.

\begin{figure*}
\centering
\includegraphics[width=1.5\columnwidth]{figures/pool-interactions.pdf}
\caption{Architecture of the SHIWA pool. Circle-terminated arrows
  indicate messages that are broadcast to all pool
  clients. Interaction \texttt{d} of Figure~\ref{fig:architectures}(IV) is
  implemented by 3 different calls to the pool: workflow
  submission (\texttt{1} \& \texttt{2}), workflow status retrieval
  (\texttt{0} \& \texttt{3}), and workflow
  retrieval (\texttt{13} \& \texttt{14}). Interaction
  \texttt{f} is implemented through 2 types of calls: workflow
  instance retrieval (\texttt{4}, \texttt{5} and \texttt{6}), and
  workflow instance update (\texttt{11} and \texttt{12}). Workflow
  instance retrieval is used by the agents to fetch work from the
  pool. Workflow instance update is used by the agents to update
  workflow statuses.  Calls \texttt{0}, \texttt{7}, \texttt{8} and
  \texttt{9} are used by
  workflow engine plugins to declare their supported language and to
  launch engines, and by workflow engines to report their status to
  the agent. These calls are specific to the SHIWA Pool implementation
  of the \texttt{agent} component and therefore have no corresponding
  representation in Figure~\ref{fig:architectures}(IV). Figure reproduced
  from~\cite{ROGE-13}.}
\label{fig:shiwa-pool-architecture}
\end{figure*}

Figure~\ref{fig:shiwa-architecture} shows the architecture
used in the SHIWA Simulation Platform for nested workflow execution
with service invocation.
\begin{figure*}
\centering
\includegraphics[width=1.5\columnwidth]{figures/shiwa-science-gateway.pdf}
\caption{Nested workflow execution through the SHIWA Science Gateway. The
  parent workflow engine is \texttt{WS-PGRADE/gUSE}, invoked as a service
  in the Science Gateway (step \texttt{1}, interaction \texttt{d} on
  Figure~\ref{fig:architectures}(V)-Right). Ten different child engines can be used by nested
  workflows, invoked through the \texttt{Submission service} (step
  \texttt{2}, interaction \texttt{d}). Each of these engines can
  submit tasks and transfer data to a distributed computing
  infrastructure (\texttt{DCI}, step \texttt{3}, interactions
  \texttt{b} and \texttt{c$_2$}). User data interactions (\texttt{c$_1$}) and
  application porting ones (\texttt{a}) are not represented. Figure
  reproduced from~\cite{terstyanszky2014enabling} with permission of
  the first author.}
\label{fig:shiwa-architecture}
\end{figure*}

\section{\correction{Metrics}}

@@

Science gateway \emph{integration complexity} is measured through the number of software components and their interactions. It breaks
down into the following 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Total components (\texttt{I$_1$}): total number of components in
  the architecture.
\item Total interactions (\texttt{I$_2$}): total number of
  interactions in the architecture.
\end{itemize}
One may wonder whether infrastructure should be counted in
\texttt{I$_1$}, since it is usually an external entity that is not
developed nor maintained by the groups who integrate workflow engines
in science gateway. We support its inclusion here because integrating
an infrastructure into the system usually requires some technical
effort (e.g., account creation, software installation, APIs, etc.),
therefore increasing complexity. \correction{We assume that software
  components contribute equally to $I_1$ and that interactions also
  contribute equally to $I_2$. In a real system, this may not be the
  case due to a range of project-specific practical reasons that
  cannot be covered in this paper. The weights given to components and
  interactions can be dynamically changed in the public spreadsheet
  (sheet ``Components and interactions'') to generate new versions of
  Table~\ref{table:evaluation} as required.} The two metrics
\correction{I$_1$ and I$_2$} are summed to obtain a global score that
measures the total number of software pieces affected by integration.

\emph{Robustness of workflow execution} measures the likelihood that
the workflow execution fails due to errors in the components or in the
interactions of the software architecture. Errors coming from the
infrastructure (e.g., unavailable data or terminated tasks) or
workflows (e.g., wrong user input or application errors) are not
covered, since they do not stem directly from the software
architecture.  Robustness is measured here as a consequence of global
complexity, since complex architectures tend to be more prone to
failure, \correction{in particular in distributed systems where
  interactions between components involve wide-area network
  communications between machines that may be operated by different
  people}. More precisely, robustness is determined by the number of
software components and interactions represented in red in the
architecture diagrams in Figure~\ref{fig:architectures}, which are
specific to workflow execution. Two metrics are used:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Specific components (\texttt{R$_1$}): number of specific
  components involved in workflow executions. \correction{For instance, the
  workflow service is a specific component because its main objective
  is to execute workflows. On the contrary, the science gateway is
  \emph{not} specific to workflow execution since it is used for a
  variety of other functions such as user authentication, data
  transfer, etc.}
\item Specific interactions (\texttt{R$_2$}): number of specific
  interactions involved in workflow executions. \correction{For
    instance, the pool-agent interaction (interaction \texttt{f}) is
    specific to workflow execution while data control (interaction
    \texttt{c$_1$}) is \emph{not} because it is used by the science gateway to transfer
    user data regardless of a particular workflow execution}.
\end{itemize}
\texttt{R$_1$} and \texttt{R$_2$} are
summed to obtain the global score for this criterion, which measures
the total number of software pieces that are specifically involved in
the workflow execution. \correction{The weights given to software components in R$_1$ and interactions in R$_2$ can be adjusted in the public spreadsheet.}

\emph{Extensibility} measures the difficulty to replace or add
elements to the architecture. It is determined by the number of interactions and
components that need to be modified when a new element is
added. Modification of a component is required when its code needs to
be updated or recompiled (science gateway or workflow service), or
when a new piece of software has to be installed (infrastructure
only). Modification of an interaction is deemed necessary when the
parameters involved in this interaction are modified.  Extensibility
breaks down into 4 metrics depending on the type of element (engine, workflow or infrastructure) that has to
be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine (\texttt{E$_1$}): number of interactions or
  components that need to be modified to integrate a new type of
  workflow engine in the architecture. Workflow engines belong to
  different types when they cannot be invoked using the same
  implementation of the interactions.
 \item Engine version upgrade (\texttt{E$_2$}): number of interactions
  or components modified to integrate a new version of a workflow
  engine in the architecture, assuming that another version of the
  same engine type is already available. Different versions of a
  workflow engine can be invoked using the same interaction
  implementation(s). When this is not the case, the different versions
  are considered as different engine types. Components are not treated
  equally regarding engine version updates. Components whose only
  function is to host the workflow engine, i.e. workflow services and
  agents, are not counted in E$_2$ because updates in such components
  are assumed to be straightforward. On the contrary, modifications of
  components that have other functions than wrapping the engine,
  i.e. science gateway and infrastructure, are counted in E$_2$
  because updates in these components require more effort, e.g. new
  release of the science gateway, gaining administrative privileges on
  the infrastructure, etc. In practice, E$_2$=0 when the workflow
  engine is wrapped in a service or in an agent, and E$_2$=1 when it is
  wrapped in another component.  
\item New workflow (\texttt{E$_3$}): number of interactions required
  to integrate a new workflow in the architecture. Adding a new
  workflow is a very common operation that does not require modifying
  software components or interactions, assuming that the engine type
  and version required to execute this workflow are already
  available. In most cases, adding a new workflow requires only
  interactions \texttt{a$_1$} and \texttt{a$_2$}, but \texttt{g}
  is required as well when workflow conversion is used. 
\item New infrastructure (\texttt{E$_4$}): number of interactions or
  components affected for the integration of a new type of infrastructure in
  the architecture. Adding a new infrastructure typically aims at providing more
  computing or storage power, enabling access to specific types of resources
  (e.g., GPUs, clouds), or  enforcing execution policies (e.g.,
  constrain data to remain in a particular network domain).
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The four metrics are summed to obtain a global score that measures the
difficulty to extend the architecture. \correction{The weights given to software components and interactions in E$_{1-4}$ can be adjusted in the public spreadsheet.} 
Note that some extensions may
affect several metrics in practice. For instance, adding a new type
of engine may help integrating new infrastructures when interactions
\texttt{b} and \texttt{c$_2$} are already present for the new engine
and the infrastructure.

>>>>

\usepackage[x11names,dvipsnames,table]{xcolor} %for use in color links
\usepackage[subrefformat=parens,labelformat=parens]{subfig}
\usepackage{svg}
\usepackage{url}
\usepackage{flushend}
\usepackage{xspace}
\usepackage{enumitem}
\usepackage[bookmarks]{hyperref}
\usepackage{multirow}
\usepackage{pbox}
\usepackage{pdfcomment}
\definecolor{ao(english)}{rgb}{0.0, 0.5, 0.0}
\newcommand{\todo}[2]{\pdfmargincomment[color=red,author=#1,open=true]{#2}}
%\newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
%\newcommand{\closednote}[4]{} % closed notes are hidden. This should obviously be a parameter of \note rather than a separate command.
%\newcommand{\answerednote}[4]{\pdfmargincomment[color=green,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}
%\newcommand{\closedanswerednote}[6]{}
\newcommand{\correction}[1]{\color{blue}#1\color{black}\xspace}

@@

\correction{The reminder of this section provides} background
information and definitions of workflow engines, science gateways and
infrastructures. In Section~\ref{sec:architectures}, we describe six
architectures that cover the main patterns that we encountered in the
various systems that we studied during the last few years. The
architectures are described within a consistent framework that
underlines the functional interactions between their main software
components. \correction{We illustrate the application of our
  architectural patterns to real systems in Section~\ref{sec:real
    systems}. In Section~\ref{sec:metrics}, we introduce metrics that
  measure integration complexity, robustness, extensibility,
  scalability and \correction{functionality}. These metrics are
  specifically designed to measure the ability of the architectures to
  address workflow-related issues commonly encountered in science
  gateways. In Section~\ref{sec:evaluation}, we evaluate the different
  architectures based on the proposed metrics.} \correction{Based on
  this evaluation, we compare and discuss the architectures in
  Section~\ref{sec:discussion} where we also highlight}
recommendations for science gateway architects and
developers. \correction{Finally, in Section~\ref{sec:application}, we
  illustrate how our evaluation framework can be used to help design
  new systems. Related work is presented in Section~\ref{sec:related}.}

@@

See Figure~\subref*{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a model in a
Ruby on Rails application\footnote{\correction{Note that the
    particular case of portlets can be a bit ambiguous: when portlets
    are deployed remotely, using protocols such as Web Services for
    Remote Portlets (WSRP), they become independent software
    components according to our definition because they could be
    operated by different teams on different hardware systems. In this
    case, the system should be classified in the service invocation
    architecture since the portlets actually become Web Services. The
    same comment holds for remote object invocation.}}. The workflow
engine and the science gateway usually share a database where
applications, users and other resources are stored.  In this model,
task and data control are both initiated from the science
gateway. Interactions \texttt{b} and \texttt{c$_2$} are initiated from
the workflow engine, while \texttt{c$_1$} comes from other parts of
the science gateway, for instance a data management interface. As in
any other model, the installation of new workflows in the science
gateway (\texttt{a$_2$}) and infrastructure (\texttt{a$_1$}) is done
by an administrator, for security reasons. This is the model adopted
in the Catania Science Gateway Framework~\cite{ardizzone2012decide}
(see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}), in the Distributed
application runtime environment
(DARE)~\cite{maddineni2012distributed}, in
Galaxy~\cite{goecks2010galaxy}, and in the LONI Pipeline
Environment~\cite{dinov2009efficient}. Note that tightly integrated
architectures may provide advanced workflow edition features which are
not covered by our analysis.
%The Catania Science Gateway Framework would be a good illustration here (workflow engine is a Java portlet), but I can't find a suitable figure about it.
%CIPRES~\cite{miller2010creating} (on which the Neuroscience gateway
%described in~\cite{sivagnanam2015early} is based) % \note{Tristan}{There is no workflow engine in there.}

@@

See Figure~\subref*{archi:sub-task}. The workflow engine is wrapped in
a particular task that can submit sub-tasks to the science gateway
through interaction \texttt{e}. The workflow engine keeps track of the
dependencies between the sub-tasks, but their execution is delegated
to the science gateway that executes them on the infrastructure
through interaction \texttt{b}. Although the science gateway has no
global vision of the workflow, it completely \correction{controls} the
sub-tasks. \correction{For instance, it may cancel} them when the
\correction{workflow} is canceled. The science gateway may also
implement mechanisms to facilitate the handling of task dependencies,
for instance \correction{lists of tasks that must be completed before a
  particular task can be executed}.

@@

\correction{\section{Architecture of real systems}}

@@

\correction{This section illustrates the architectural patterns on a
  few real systems. The architecture diagrams presented are the
  original ones as published by the system developers, which we
  annotated using the interactions and components in
  Figure~\ref{fig:architectures}, in red color. Application porting
  interactions \texttt{a$_1$} and \texttt{a$_2$} are not represented
  on these diagrams.

The architecture of the Catania Science Gateway Framework (CSGF) is
shown in Figure~\ref{fig:csgf}. In CSGF, workflows are integrated as
Liferay portlets, i.e., as part of the Science Gateway when no
specific protocol is used for remote invocation. A specific portlet
called \texttt{mi-parallel-portlet} allows to create job collections,
parametric jobs and split-and-merge types of workflows. The science
gateway also includes a set of services to manage data and tasks
(jobs) on a wide array of grid middleware systems through the SAGA
API.
\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/CSGF}
\caption{\correction{Architecture of the Catania Science Gateway Framework (tight
  integration). Figure adapted from~\cite{ardizzone2012decide} with
  permission of the first author} \todo{approval still pending}.}
\label{fig:csgf}
\end{figure}

@@

Figure~\ref{fig:vip-architecture} shows the architecture of the
Virtual Imaging Platform, where the MOTEUR workflow
engine~\cite{GLAT-08i} is invoked through a service. The interactions
in the VIP architecture map to the ones defined by our framework in a
straightforward manner, except for interaction \texttt{c$_2$}
(management of workflow data) which, in VIP, is triggered by the
workflow tasks instead of the workflow engine. However, this does not
contradict the service invocation architecture since the tasks are
generated by the workflow engine.
\begin{figure*}
%\def\svgwidth{1.5\columnwidth}
\includegraphics[width=0.9\textwidth]{figures/VIP.eps}
\caption{Architecture of the Virtual Imaging Platform (service
  invocation).  Step \texttt{3} corresponds to interaction \texttt{d}
  in Figure~\ref{fig:architectures}(II), step \texttt{2} corresponds
  to interaction \texttt{c$_1$}, step \texttt{4} maps to interaction
  \texttt{b}, and steps 7 and 9 map to \texttt{c$_2$} (in VIP,
  workflow data transfers are performed by the tasks and embedded in
  their descriptions).}
\label{fig:vip-architecture}
\end{figure*}

@@

Task encapsulation is used in the CBRAIN gateway~\cite{SHER-14} to
integrate the FSL toolkit~\cite{Jenkinson2012782} and the PSOM
workflow engine~\cite{bellec2012pipeline}. The CBRAIN-FSL integration,
shown in Figure~\subref*{fig:cbrain-fsl-architecture}, allows
leveraging FSL pipelines written in low-level workflow languages
(Linux executables and scripts) that submit tasks uniformly through a
specific tool called fsl\_sub. The science gateway is actually split
in two services, the CBRAIN portal and the CBRAIN execution server,
deployed on a head node of the infrastructure. The mapping of
interactions \texttt{b}, \texttt{c$_1$} and \texttt{e} to the
CBRAIN-FSL interactions is straightforward.

@@

\begin{figure*}
\centering
\subfloat[CBRAIN-FSL integration.]{
\includegraphics[width=\columnwidth]{figures/CBRAIN-FSL.eps}
\label{fig:cbrain-fsl-architecture}
} \hfill \subfloat[CBRAIN-PSOM integration. Figure
reproduced from~\cite{GLAT-16}.]{
  \includegraphics[width=\columnwidth]{figures/CBRAIN-PSOM.eps}
\label{fig:cbrain-psom-architecture}
}
\caption{CBRAIN architecture for workflow engine integration
  (task encapsulation).  1: User sends data and workflow execution request to
  storage server(s) and CBRAIN portal. 2: CBRAIN portal sends
  execution request to execution server on cluster. 3: Execution
  server transfers data from storage server(s). 4-5: Execution server
  starts workflow engine (FSL tool or PSOM master) via task
  scheduler. 6: Workflow engine submits sub-tasks to execution server
  (FSL tasks or PSOM agents). 7-8: Execution server starts sub-tasks
  through task scheduler (FSL sub-tasks or PSOM workers). FSL sub-tasks will run locally
  instead of being submitted again to CBRAIN through interaction 6. 8':
  (PSOM only) PSOM master and workers execute workflow. 9. Execution
  server transfers results to storage server(s). Interaction
  \texttt{b} in Figure~\ref{fig:architectures}(III) is implemented by steps
  \texttt{4}, \texttt{5}, \texttt{7} and \texttt{8} (regular
  interactions with batch manager). Interaction \texttt{c} is
  implemented by steps \texttt{3} and \texttt{9}. Interaction
  \texttt{e} is implemented by step \texttt{6} (for FSL: through a
  modified version of the fsl\_sub script available in
  \url{https://github.com/aces/cbrain-plugins-neuro}; for PSOM:
  through a specific development in PSOM).}
\label{fig:cbrain-sub-tasking}
\end{figure*}

The pool model was implemented in the SHIWA pool~\cite{ROGE-13}
diagrammed in Figure~\ref{fig:shiwa-pool-architecture}, where agents
can wrap different types of workflow engines to execute workflows
expressed with different languages. The software components map
directly to the ones in our architecture, except that the
infrastructure is not represented. The interactions described in the
SHIWA pool are much finer-grained than in our model, but they can be
grouped to match interactions \texttt{d} and \texttt{f}, as explained
in the Figure caption.

\begin{figure*}
\centering
\includegraphics[width=1.5\columnwidth]{figures/pool-interactions.pdf}
\caption{Architecture of the SHIWA pool. Circle-terminated arrows
  indicate messages that are broadcast to all pool
  clients. Interaction \texttt{d} of Figure~\ref{fig:architectures}(IV) is
  implemented by 3 different calls to the pool: workflow
  submission (\texttt{1} \& \texttt{2}), workflow status retrieval
  (\texttt{0} \& \texttt{3}), and workflow
  retrieval (\texttt{13} \& \texttt{14}). Interaction
  \texttt{f} is implemented through 2 types of calls: workflow
  instance retrieval (\texttt{4}, \texttt{5} and \texttt{6}), and
  workflow instance update (\texttt{11} and \texttt{12}). Workflow
  instance retrieval is used by the agents to fetch work from the
  pool. Workflow instance update is used by the agents to update
  workflow statuses.  Calls \texttt{0}, \texttt{7}, \texttt{8} and
  \texttt{9} are used by
  workflow engine plugins to declare their supported language and to
  launch engines, and by workflow engines to report their status to
  the agent. These calls are specific to the SHIWA Pool implementation
  of the \texttt{agent} component and therefore have no corresponding
  representation in Figure~\ref{fig:architectures}(IV). Figure adapted
  from~\cite{ROGE-13}.}
\label{fig:shiwa-pool-architecture}
\end{figure*}

Figure~\ref{fig:shiwa-architecture} shows the architecture used in the
SHIWA Simulation Platform for nested workflow execution with service
invocation. The parent workflow engine is \texttt{WS-PGRADE/gUSE},
invoked as a service in the Science Gateway (interaction
\texttt{d}). Ten different child engines can be used by nested
workflows, invoked through the \texttt{Submission service}
(interaction \texttt{d}). Each of these engines can submit tasks and
transfer data to a distributed computing infrastructure (\texttt{DCI}
interactions \texttt{b} and \texttt{c$_2$}). User data interactions
(\texttt{c$_1$}) and interactions between the parent engine and the
infrastructure (\texttt{b} and \texttt{c$_2$}) are not represented.
\begin{figure*}
\centering
\includegraphics[width=1.5\columnwidth]{figures/shiwa-science-gateway.png}
\caption{Nested workflow execution through the SHIWA Science Gateway. Figure
  adapted from~\cite{terstyanszky2014enabling} with permission of
  the first author.}
\label{fig:shiwa-architecture}
\end{figure*}

Figure~\ref{fig:tavaxy} shows the architecture of Tavaxy where
workflows in various formats (tSCUFL, SCUFL, t2flow and JSON) can be
imported. As detailed in~\cite{Abouelhoda2012}, workflow conversion
(interaction \texttt{g}) is implemented through an interactive
authoring module where users can check and edit workflow imports, and
a mapper module that performs some language substitutions. The
workflow engine is a re-engineered version of the Galaxy engine that
also includes some components of Taverna.
\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/Tavaxy.eps}
\caption{\correction{Architecture of Tavaxy (workflow conversion). Figure adapted
  from~\cite{Abouelhoda2012} with permission of the first author} \todo{approval is still pending}.}
\label{fig:tavaxy}
\end{figure}

}

\section{\correction{Evaluation metrics}}

@@

\subsection{\correction{Integration complexity}}

\correction{Integration complexity} is measured through the number of
software components and their interactions. It breaks down into the
following 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Total components (\texttt{I$_1$}): total number of components in
  the architecture.
\item Total interactions (\texttt{I$_2$}): total number of
  interactions in the architecture.
\end{itemize}
One may wonder whether infrastructure should be counted in
\texttt{I$_1$}, since it is usually an external entity that is not
developed nor maintained by the groups who integrate workflow engines
in science gateway. We support its inclusion here because integrating
an infrastructure into the system usually requires some technical
effort (e.g., account creation, software installation, APIs, etc.),
therefore increasing complexity. \correction{We assume that software
  components contribute equally to $I_1$ and that interactions also
  contribute equally to $I_2$. In a real system, this may not be the
  case due to a range of project-specific practical reasons that
  cannot be covered in this paper. The weights given to components and
  interactions can be dynamically changed in the public spreadsheet
  (sheet ``Components and interactions'') to generate new versions of
  Table~\ref{table:evaluation} as required.} The two metrics
\correction{I$_1$ and I$_2$} are summed to obtain a global score that
measures the total number of software pieces affected by integration.

\subsection{\correction{Robustness of workflow execution}}

Robustness of workflow execution measures the likelihood that
the workflow execution fails due to errors in the components or in the
interactions of the software architecture. Errors coming from the
infrastructure (e.g., unavailable data or terminated tasks) or
workflows (e.g., wrong user input or application errors) are not
covered, since they do not stem directly from the software
architecture.  Robustness is measured here as a consequence of global
complexity, since complex architectures tend to be more prone to
failure, \correction{in particular in distributed systems where
  interactions between components involve wide-area network
  communications between machines that may be operated by different
  people}. More precisely, robustness is determined by the number of
software components and interactions represented in red in the
architecture diagrams in Figure~\ref{fig:architectures}, which are
specific to workflow execution. Two metrics are used:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Specific components (\texttt{R$_1$}): number of specific
  components involved in workflow executions. \correction{For instance, the
  workflow service is a specific component because its main objective
  is to execute workflows. On the contrary, the science gateway is
  \emph{not} specific to workflow execution since it is used for a
  variety of other functions such as user authentication, data
  transfer, etc.}
\item Specific interactions (\texttt{R$_2$}): number of specific
  interactions involved in workflow executions. \correction{For
    instance, the pool-agent interaction (interaction \texttt{f}) is
    specific to workflow execution while data control (interaction
    \texttt{c$_1$}) is \emph{not} because it is used by the science gateway to transfer
    user data regardless of a particular workflow execution}.
\end{itemize}
\texttt{R$_1$} and \texttt{R$_2$} are
summed to obtain the global score for this criterion, which measures
the total number of software pieces that are specifically involved in
the workflow execution. \correction{The weights given to software components in R$_1$ and interactions in R$_2$ can be adjusted in the public spreadsheet.}

\subsection{\correction{Extensibility}}

Extensibility measures the difficulty to replace or add elements to
the architecture. It is determined by the number of interactions and
components that need to be modified when a new element is
added. Modification of a component is required when its code needs to
be updated or recompiled (science gateway or workflow service), or
when a new piece of software has to be installed (infrastructure
only). Modification of an interaction is deemed necessary when the
parameters involved in this interaction are modified.  Extensibility
breaks down into 4 metrics depending on the type of element (engine,
workflow or infrastructure) that has to be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine (\texttt{E$_1$}): number of interactions or
  components that need to be modified to integrate a new type of
  workflow engine in the architecture. Workflow engines belong to
  different types when they cannot be invoked using the same
  implementation of the interactions.
 \item Engine version upgrade (\texttt{E$_2$}): number of interactions
  or components modified to integrate a new version of a workflow
  engine in the architecture, assuming that another version of the
  same engine type is already available. Different versions of a
  workflow engine can be invoked using the same interaction
  implementation(s). When this is not the case, the different versions
  are considered as different engine types. Components are not treated
  equally regarding engine version updates. Components whose only
  function is to host the workflow engine, i.e. workflow services and
  agents, are not counted in E$_2$ because updates in such components
  are assumed to be straightforward. On the contrary, modifications of
  components that have other functions than wrapping the engine,
  i.e. science gateway and infrastructure, are counted in E$_2$
  because updates in these components require more effort, e.g. new
  release of the science gateway, gaining administrative privileges on
  the infrastructure, etc. In practice, E$_2$=0 when the workflow
  engine is wrapped in a service or in an agent, and E$_2$=1 when it is
  wrapped in another component.  
\item New workflow (\texttt{E$_3$}): number of interactions required
  to integrate a new workflow in the architecture. Adding a new
  workflow is a very common operation that does not require modifying
  software components or interactions, assuming that the engine type
  and version required to execute this workflow are already
  available. In most cases, adding a new workflow requires only
  interactions \texttt{a$_1$} and \texttt{a$_2$}, but \texttt{g}
  is required as well when workflow conversion is used. 
\item New infrastructure (\texttt{E$_4$}): number of interactions or
  components affected for the integration of a new type of infrastructure in
  the architecture. Adding a new infrastructure typically aims at providing more
  computing or storage power, enabling access to specific types of resources
  (e.g., GPUs, clouds), or  enforcing execution policies (e.g.,
  constrain data to remain in a particular network domain).
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The four metrics are summed to obtain a global score that measures the
difficulty to extend the architecture. \correction{The weights given to software components and interactions in E$_{1-4}$ can be adjusted in the public spreadsheet.} 
Note that some extensions may
affect several metrics in practice. For instance, adding a new type
of engine may help integrating new infrastructures when interactions
\texttt{b} and \texttt{c$_2$} are already present for the new engine
and the infrastructure.

\subsection{\correction{Scalability}}
